(Deep) Induction Rules for GADTs\vspace*{-0.1in}

### 2021-08-29 induction-paper/revised-induction-paper.tex
revised induction paper 8-29

\begin{abstract} 
  Deep data types are those that are contructed from other data types,
  including, possibly, themselves.  In this case, they are said to be
  truly nested.  Deep induction is an extension of structural
  induction that traverses {\em all} of the structure in a deep data
  type, propagating predicates on its primitive data throughout the
  entire structure.  Deep induction can be used to prove properties of
  nested types, including truly nested types, that cannot be proved
  via structural induction.  In this paper we show how to extend deep
  induction to deep GADTs that are not truly nested GADTs. We also
  show that the techniques developed in this paper do not suffice for
  extending deep induction to truly nested GADTs.\looseness=-1
\end{abstract}

@@

\[\begin{array}{l}
\mathsf{\forall\, (A : Set)\,(P : Rose\,A \to Set) \to P\,empty \to}\\
\mathsf{\hspace*{0.15in}
 (\forall\, (a : A)\,(ts :
  List\,(Rose\,A)) \to P\,(node\,a\,ts)) \to 
  \forall \,(x : Rose\,A) \to P\,x}
\end{array}\]
\noindent
Unfortunately, this is neither the induction rule we intuitively
expect, nor is it expressive enough to prove even basic properties of
rose trees that ought to be amenable to inductive proof. What is
needed here is an enhanced notion of induction that, when specialized
to rose trees, will propagate the predicate $\mathsf{P}$ through the
outer list structure and to the rose trees sitting inside
$\mathsf{node}$'s list argument. More generally, this enhanced notion
of induction should traverse {\em all} of the layers present in a data
structure, propagating suitable predicates to {\em all} of the data it
contains. With data types becoming ever more advanced, and with deeply
structured such types becoming increasingly ubiquitous in
formalizations, it is critically important that proof assistants
%{\color{red} (e.g., Coq, Agda, Lean)}
are able to automatically generate genuinely useful induction rules
for data types that go well beyond traditional ADTs. Such data types
include not just deep ADTs, but also (truly\footnote{A truly nested
  type is a nested type that is defined over itself. The data type
  $\mathsf{Bush}$ in Section~\ref{sec:ADTs-and-nesteds} provides a
  concrete example.}) nested types~\cite{bm98},
%, such as $\mathsf{Rose}$ above, as well as truly nested types
generalized algebraic data types
(GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
families~\cite{dyb94}, and deep variants of all of these.

@@

Deep induction was shown in~\cite{jp20} to deliver induction rules
appropriate to deep nested types, including deep ADTs.  For the deep
ADT of rose trees, for example, it gives the following genuinely
useful induction rule:

@@

\noindent
%Here, $\mathsf{List^\land}$ lifts its predicate argument $\mathsf{P}$
%on data of type $\mathsf{Rose\,a}$ to a predicate on data of type
%$\mathsf{List\,(Rose\,a)}$ asserting that $\mathsf{P}$ holds for every
%element of its argument list, and similarly for the lifting
%$\mathsf{Rose^\land}$ and the predicate $\mathsf{Q}$.
Here, $\mathsf{List^\land}$ (resp., $\mathsf{Rose^\land}$) lifts its
predicate argument $\mathsf{P}$ (resp., $\mathsf{Q}$) on data of type
$\mathsf{Rose\,A}$ (resp., $\mathsf{A}$) to a predicate on data of
type $\mathsf{List\,(Rose\,A)}$ (resp., $\mathsf{Rose\,A}$) asserting
that $\mathsf{P}$ (resp., $\mathsf{Q}$) holds for every element of its
list (resp., rose tree) argument.\footnote{Predicate liftings such as
  $\mathsf{List^\land}$ and $\mathsf{Rose^\land}$ can either be
  supplied as primitives or generated automatically from their
  associated data type definitions as described in
  Section~\ref{sec:ADTs-and-nesteds} below. The predicate lifting for
  a container type like $\mathsf{List\,A}$ or $\mathsf{Rose\,A}$
  simply traverses containers of that type and applies its predicate
  argument pointwise to the constituent data of type $\mathsf{A}$.
  The ability to define predicate liftings for more general data types
  will be critical to deriving their deep induction rules in
  Section~\ref{sec:framework}.}

Deep induction was also shown in~\cite{jp20} to
% be the missing piece making it possible to
%be the theretofore missing component needed to
% be the essential ingredient making it possible to
deliver the first-ever induction rules --- structural or otherwise ---
for the $\mathsf{Bush}$ data type~\cite{bm98} and other truly nested
types. Deep induction for ADTs and (truly) nested types is reviewed in
Section~\ref{sec:ADTs-and-nesteds} below.

This paper shows how to extend deep induction to proper GADTs, i.e.,
GADTs that go beyond nested types (and thus also beyond ADTs).  A
constructor for such a GADT $\mathsf{G}$ may, like a constructor for a
nested type, take as arguments data whose types involve instances of
$\mathsf{G}$ other than the one being defined. These can even include
instances involving $\mathsf{G}$ itself. If $\mathsf{G}$ is a proper
GADT then at least one of its constructors will also have such a
structured instance of $\mathsf{G}$ --- albeit one not involving
$\mathsf{G}$ itself --- as its codomain.
%But while the types of the arguments to a GADT's constructor can
%include instances of $\mathsf{G}$ that themselves involve
%$\mathsf{G}$, its return type cannot.
For example, the constructor $\mathsf{pair}$ for the GADT

@@

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences\footnote{The type of $\mathsf{Seq}$ is actually
  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
  explicit tracking of Agda universe levels in this paper.} only
constructs sequences of pairs, rather than sequences of arbitrary
type, as does $\mathsf{const}$. If one or more of the constructors for
a GADT $\mathsf{G}$ return structured instances of $\mathsf{G}$, then
the GADT will have two distinct, but equally natural, semantics: a
functorial semantics interpreting it as a left Kan
extension~\cite{mac71}, and a parametric semantics interpreting it as
the interpretation of its Church encoding~\cite{atk12,vw10}. As
detailed in~\cite{jgj21}, a key difference in the two semantics is
that the former views GADTs as their {\em functorial
  completions}~\cite{jp19}, and thus as containing more data than just
those expressible in syntax. By contrast, the latter views them as
what might be called {\em syntax-only} GADTs. Happily, these two views
of GADTs coincide for those GADTs that are ADTs or other nested
types. However, both they and their attendant properties differ
greatly for proper GADTs. In fact, the functorial and parametric
semantics for proper GADTs are sufficiently disparate that, by
contrast with the semantics customarily given for ADTs and other
nested types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how
to define a functorial parametric semantics for them~\cite{jgj21}.

This observation seems, at first, to be a death knell for the prospect
of extending deep induction to GADTs. Indeed, induction can be seen as
unary parametricity, so GADTs viewed as their functorial completions
do not obviously support induction rules.  This makes sense
intuitively: induction is a syntactic proof technique, so it may not
be possible to use it to prove properties of those elements of a
GADT's functorial completion that are not expressible in syntax. All
is not lost, however. As we show below, the syntax-only view of GADTs
determined by their Church encodings {\em does} support induction
rules --- including deep induction rules --- for GADTs.  Indeed, this
paper gives the first-ever induction rules --- deep or otherwise ---
for proper GADTs. But it actually delivers far more: it gives a
general framework for deriving deep induction rules for deep GADTs
directly from their syntax. This framework can serve as a basis for
extending modern proof assistants' automatic generation of structural
induction rules for ADTs to automatic generation of deep induction
rules for GADTs. In addition, as for ADTs and other nested types, the
structural induction rule for any GADT can be recovered from its deep
induction rule simply by taking the custom predicates in its deep
induction rule to be constantly $\mathsf{True}$-valued (i.e.,
constantly $\mathsf{\top}$-valued) predicates.

Significantly, deep induction rules for GADTs cannot be derived by
somehow extending the approach of~\cite{jp20} to syntax-only
GADTs. Indeed, the approach taken there makes crucial use of the
functoriality of data types' interpretations from~\cite{jp19}, and
functoriality is precisely what interpreting GADTs as the
interpretations of their Church encodings fails to deliver. Our
approach is to instead first give a predicate lifting styled after
those of~\cite{jp20}, together with a (deep) induction rule, for the
simplest --- and arguably most important --- GADT, namely the equality
GADT. (See Section~\ref{sec:ind-equal}.)  We then derive the deep
induction rule for a more complex GADT $\mathsf{G}$ by {\em i}) using
the equality GADT to represent $\mathsf{G}$ as its so-called {\em
  Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em
  ii}) using the predicate liftings for the equality GADT and the
other GADTs appearing in the definition of $\mathsf{G}$ to
appropriately thread the custom predicates for the primitive types
appearing in $\mathsf{G}$ throughout $\mathsf{G}$'s structure. This
two-step process delivers deep induction rules for a very general
class of deep GADTs. In Section~\ref{sec:GADTs} we introduce a series
of increasingly complex GADTs as running examples, and in
Section~\ref{sec:deep-ind-GADTs} we derive a deep induction rule for
each of them. In particular, we derive the deep induction rule for the
$\mathsf{Seq}$ data type in~\eqref{eq:seq} in
Section~\ref{sec:ind-seq}. We present our general framework for
deriving (deep) induction rules for (deep) GADTs in
Section~\ref{sec:framework}, and observe that the derivations in
Section~\ref{sec:deep-ind-GADTs} are all instances of it.  In
Section~\ref{sec:GADT-nested} we show that, by contrast with truly
nested types, which do have a functorial semantics, syntax-only GADTs'
lack of functoriality means that it is not clear how to extend
induction --- deep or otherwise --- to {\em truly nested GADTs}, i.e.,
to proper GADTs whose recursive occurrences appear below
themselves.\footnote{Note carefully the distinction between a GADT
  that is not a nested type --- i.e., a proper GADT --- and a proper
  GADT that is not a truly nested GADT.  In fact, truly nested types
  are not proper GADTs and truly nested GADTs are not (truly) nested
  types. Confusion is understandable in light of the original, and now
  well-established, use of the term ``nested type'' to refer to types
  that do not necessarily involve actual nesting~\cite{bm98}.}  This
does not appear to be much of a restriction, however, since truly
nested GADTs do not, to our knowledge, appear in practice or in the
literature. Section~\ref{sec:app} comprises a case study in using deep
induction. All of the deep induction rules appearing in this paper
have been derived by instantiating our general framework. Our Agda
code implementing them is available at~\cite{web-page}.

@@

{\bf Additional Related Work\/} Various techniques for deriving
induction rules for data types that go beyond ADTs have been
studied. For example, Fu and Selinger~\cite{fs18} show, via examples,
how to derive induction rules for arbitrary nested
types. Unfortunately, however, their technique is rather {\em ad hoc},
so is unclear how to generalize it to nested types other than the
specific ones studied there. Moreover,~\cite{fs18} actually derives
induction rules for data types {\em related} to the original nested
types rather than for the original nested types themselves, and it is
unclear whether or not the derived rules are sufficiently expressive
to prove all results about the original nested types that we would
expect to be provable by induction. This latter point echoes the issue
with Coq-derived induction rule for rose trees raised in
Section~\ref{sec:intro}, which has the unfortunate effect of forcing
users to manually write induction (and other) rules for such types for
use in that system. Tassi~\cite{tas19} derives induction rules for
data type definitions in Coq using unary parametricity. His technique
seems to be essentially equivalent to that of~\cite{jp19} for nested
types, although he does not permit true nesting. More recently,
Ullrich~\cite{ull20} has implemented a plugin in MetaCoq to generate
induction rules for nested types. This plugin is also based on unary
parametricity, and true nesting still is not permitted.  As far as we
know, no attempts have been made to extend either implementation to
truly nested types or to proper GADTs or their deep variants. In fact,
we know of no work other than that reported here that specifically
addresses induction rules for such data types.

@@

Under some circumstances deep induction can be mimicked by
hand-threading applications of structural induction through the layers
of data comprising a deep data type. But this is not the case if,
e.g., one or more of the custom predicates in the data type's deep
induction rule cannot be defined by structural induction.

@@

GADTs are used in precisely those situations in which different
behaviors at different instances of data types are desired. This is
achieved by allowing the programmer to give the type signatures of the
GADT's data constructors independently, and then taking advantage of
pattern matching to force the desired type refinement. For example,
the {\em equality} GADT
\begin{equation}\label{eq:equal}
\begin{array}{l}
\mathsf{data\ Equal : Set \to Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,refl :\, \forall \{A : Set\} \to Equal\,A\,A}
\end{array}
\end{equation}
is parameterized by two type indices, but it is only possible to
construct data elements of type $\mathsf{Equal\,A\,B}$ if $\mathsf{A}$
and $\mathsf{B}$ are instantiated at the same type. If the types
$\mathsf{A}$ and $\mathsf{B}$ are syntactically identical then the
type $\mathsf{Equal\,A\,B}$ contains the single data element
$\mathsf{refl}$. It contains no data elements otherwise.

@@

The equality constraints engendered by GADTs' data constructors makes
deriving (deep) induction rules for them more involved than for ADTs
and other nested types. Nevertheless, we show in this section how to
do so. We first illustrate the key components of our approach by
deriving deep induction rules for the three specific GADTs introduced
in Section~\ref{sec:GADTs}. Then, in Section~\ref{sec:framework}, we
abstract these to a general framework that can be applied to any deep
GADT that is not a truly nested GADT. As hinted above, the predicate
lifting for the equality GADT plays a central role in deriving both
structural and deep induction rules for more general GADTs.

@@

To define the (deep) induction rule for any (deep) GADT $\mathsf{G}$
we first need to define a predicate lifting that maps a predicate on
a type $\mathsf{A}$ to a predicate on $\mathsf{G\,A}$. Such a predicate
lifting $\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set)
  \to (B \to Set) \to Equal\,A\,B \to Set}$ for $\mathsf{Equal}$ is
defined by $\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a :
  A) \to Equal\,(Q\,a)(Q'\,a)}$.  It takes two predicates on the same
type as input and tests them for extensional equality.
%
Next, we need to associate with each data constructor $\mathsf{c}$ of
$\mathsf{G}$ an {\em induction hypothesis} asserting that, if the
custom predicate arguments to a predicate $\mathsf{P}$ on $\mathsf{G}$
can be lifted to $\mathsf{G}$ itself, then $\mathsf{c}$ {\em respects}
$\mathsf{P}$, i.e., $\mathsf{c}$ constructs data satisfying the
instance of $\mathsf{P}$ at those custom predicates. The following
induction hypothesis $\mathsf{dIndRefl}$ is thus associated with the
$\mathsf{refl}$ constructor for $\mathsf{Equal}$:
\begin{equation*}\label{eq:ind-refl}
\begin{array}{l}
\mathsf{\lambda (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
  Set) \to Equal\,A\,B \to Set) \to} \\ 
\quad\mathsf{\forall (C : Set) (Q\, Q' : C \to Set) \to
  Equal^{\wedge}\,C\,C\,Q\,Q'\,refl \to P\,C\,C\,Q\,Q'\,refl} 
\end{array}
\end{equation*}

The deep induction rule for $\mathsf{G}$ now states that, if all of
$\mathsf{G}$'s data constructors respect a predicate $\mathsf{P}$,
then $\mathsf{P}$ is satisfied by every element of $\mathsf{G}$ to
which the custom predicate arguments to $\mathsf{P}$ can be
successfully lifted.  The deep induction rule for $\mathsf{Equal}$ is
thus
\begin{equation}\label{eq:ind-equal}
\begin{array}{l}
\mathsf{\forall (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
  Set) \to Equal\,A\,B \to Set) \to dIndRefl\,P \to}\\ \quad 
  \mathsf{\forall (A\,B : Set) (Q_A : A \to Set) (Q_B : B \to Set) (e:
  Equal\,A\,B) \to Equal^{\wedge}\,A\,B\,Q_A\,Q_B\,e \to
  P\,A\,B\,Q_A\,Q_B\,e}
\end{array}
\end{equation}

@@

We can generalize the approach taken in
Section~\ref{sec:deep-ind-GADTs} to a general framework for deriving
deep induction rules for deep GADTs that are not truly nested
GADTs. We will treat deep GADTs of the form

@@

\noindent
This grammar is subject to the following restrictions. In the third
clause the type constructor $\mathsf{F_1}$ does not contain
$\mathsf{G}$. In the fourth clause, none of the $\mathsf{\alpha}$-many
type constructors in $\mathsf{\ol{F_1}}$ contains $\mathsf{G}$. This
prevents nesting, which would make it impossible to give an induction
rule for $\mathsf{G}$; see Section~\ref{sec:GADT-nested} below. In the
fifth and sixth clauses, $\mathsf{H : Set^\gamma \to Set}$ is the
syntactic reflection of some functor, and thus has an associated map
function. It is worth noting that the fifth clause subsumes the cases
in which $\mathsf{F\,G\,\ol{B}}$ is a closed type or one of the
$\mathsf{B_i}$, and that $\mathsf{H}$ can be the data type constructor
for any (truly) nested type. From the map function for $\mathsf{H}$ we
can also construct a map function

>>>>

\begin{abstract} 
  Deep data types are those that are contructed from other data types,
  including, possibly, themselves.  In this case, they are said to be
  truly nested.  Deep induction is an extension of structural
  induction that traverses {\em all} of the structure in a deep data
  type, propagating predicates on its primitive data throughout the
  entire structure.  Deep induction can be used to prove properties of
  nested types, including truly nested types, that cannot be proved
  via structural induction.  In this paper we show how to extend deep
  induction to GADTs that are not truly nested GADTs. We also show
  that the techniques developed in this paper do not suffice for
  extending deep induction to truly nested GADTs.\looseness=-1
\end{abstract}

@@

\[\begin{array}{l}
\mathsf{\forall\, (A : Set)\,(P : Rose\,A \to Set) \to P\,empty \to}\\
\mathsf{\hspace*{0.15in}
 (\forall\, (a : A)\,(ts :
  List\,(Rose\,A)) \to P\,(node\,a\,ts)) \to 
  \forall \,(x : Rose\,A) \to P\,x}
\end{array}\]
\noindent
Unfortunately, this is neither the induction rule we intuitively
expect, nor is it expressive enough to prove even basic properties of
rose trees that ought to be amenable to inductive proof. What is
needed here is an enhanced notion of induction that, when specialized
to rose trees, will propagate the predicate $\mathsf{P}$ through the
outer list structure and to the rose trees sitting inside
$\mathsf{node}$'s list argument. More generally, this enhanced notion
of induction should traverse {\em all} of the layers present in a data
structure, propagating suitable predicates to {\em all} of the data it
contains.  With data types becoming ever more advanced, and with
deeply structured such types becoming increasingly ubiquitous in
formalizations, such an enhanced notion of induction is essential if
proof assistants are to be
%it is critically important that proof assistants
%%{\color{red} (e.g., Coq, Agda, Lean)}
%are
able to automatically generate genuinely useful induction rules for
data types that go
%well
beyond traditional ADTs. These
%Such data types
include not just deep ADTs, but also (truly\footnote{A truly nested
  type is a nested type that is defined over itself. The data type
  $\mathsf{Bush}$ in Section~\ref{sec:ADTs-and-nesteds} provides a
  concrete example.}) nested types~\cite{bm98},
%, such as $\mathsf{Rose}$ above, as well as truly nested types
generalized algebraic data types
(GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
families~\cite{dyb94}, and deep variants of all of these.

@@

Deep induction was shown in~\cite{jp20} to deliver induction rules
appropriate to nested types, including ADTs.  For the (deep) ADT of
rose trees, for example, it gives the following genuinely useful
induction rule:

@@

\noindent
%Here, $\mathsf{List^\land}$ lifts its predicate argument $\mathsf{P}$
%on data of type $\mathsf{Rose\,a}$ to a predicate on data of type
%$\mathsf{List\,(Rose\,a)}$ asserting that $\mathsf{P}$ holds for every
%element of its argument list, and similarly for the lifting
%$\mathsf{Rose^\land}$ and the predicate $\mathsf{Q}$.
Here, $\mathsf{List^\land}$ (resp., $\mathsf{Rose^\land}$) lifts its
predicate argument $\mathsf{P}$ (resp., $\mathsf{Q}$) on data of type
$\mathsf{Rose\,A}$ (resp., $\mathsf{A}$) to a predicate on data of
type $\mathsf{List\,(Rose\,A)}$ (resp., $\mathsf{Rose\,A}$) asserting
that $\mathsf{P}$ (resp., $\mathsf{Q}$) holds for every element of its
list (resp., rose tree) argument.\footnote{Predicate liftings such as
  $\mathsf{List^\land}$ and $\mathsf{Rose^\land}$ can either be
  supplied as primitives or generated automatically from their
  associated data type definitions as described in
  Section~\ref{sec:ADTs-and-nesteds} below. The predicate lifting for
  a container type like $\mathsf{List\,A}$ or $\mathsf{Rose\,A}$
  simply traverses containers of that type and applies its predicate
  argument pointwise to the constituent data of type $\mathsf{A}$.
  The ability to define predicate liftings for more general data types
  will be critical to deriving their deep induction rules in
  Section~\ref{sec:framework}.} Deep induction was also shown
in~\cite{jp20} to
% be the missing piece making it possible to
%be the theretofore missing component needed to
% be the essential ingredient making it possible to
deliver the first-ever induction rules --- structural or otherwise ---
for the $\mathsf{Bush}$ data type~\cite{bm98} and other truly nested
types. Deep induction for ADTs and (truly) nested types is reviewed in
Section~\ref{sec:ADTs-and-nesteds} below.

This paper shows how to extend deep induction to proper GADTs, i.e.,
GADTs that are not nested types (and thus are not ADTs).  A
constructor for a GADT $\mathsf{G}$ may, like a constructor for a
nested type, take as arguments data whose types involve instances of
$\mathsf{G}$ other than the one being defined. These can even include
instances involving $\mathsf{G}$ itself. But if $\mathsf{G}$ is a
proper GADT, then at least one of its constructors will also have a
structured instance of $\mathsf{G}$ --- albeit one not involving
$\mathsf{G}$ itself --- as its codomain.
%But while the types of the arguments to a GADT's constructor can
%include instances of $\mathsf{G}$ that themselves involve
%$\mathsf{G}$, its return type cannot.
For example, the constructor $\mathsf{pair}$ for the GADT

@@

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences\footnote{The type of $\mathsf{Seq}$ is actually
  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
  explicit tracking of Agda universe levels in this paper.} only
constructs sequences of data whose types are pair-structured, rather
than sequences of arbitrary type, as does $\mathsf{const}$.

If one or more of the constructors for a GADT $\mathsf{G}$ return
structured instances of $\mathsf{G}$, then the GADT will have two
distinct, but equally natural, semantics: a functorial semantics
interpreting it as a left Kan extension~\cite{mac71}, and a parametric
semantics interpreting it as the interpretation of its Church
encoding~\cite{atk12,vw10}. As detailed in~\cite{jgj21}, a key
difference in the two semantics is that the former views GADTs as
their {\em functorial completions}~\cite{jp19}, and thus as containing
more data than just those expressible in syntax. By contrast, the
latter views them as what might be called {\em syntax-only}
GADTs. Happily, these two views of GADTs coincide for those GADTs that
are ADTs or (other, including truly) nested types.  However, both they
and their attendant properties differ greatly for proper GADTs. In
fact, the functorial and parametric semantics for proper GADTs are
sufficiently disparate that, by contrast with the semantics
customarily given for ADTs and nested
types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how to
define a functorial parametric semantics for GADTs~\cite{jgj21}.

This observation seems, at first, to be a death knell for the prospect
of extending deep induction to GADTs. Indeed, induction can be seen as
unary parametricity, so GADTs viewed as their functorial completions
do not obviously support induction rules.  This makes sense
intuitively: induction is a syntactic proof technique, so it may not
be possible to use it to prove properties of those elements of a
GADT's functorial completion that are not expressible in syntax. All
is not lost, however. As we show below, the syntax-only view of GADTs
determined by their Church encodings {\em does} support induction
rules --- including deep induction rules --- for GADTs. Indeed, this
paper gives the first-ever deep induction rules
%--- deep or otherwise ---
for proper GADTs.
%It also shows how the structural induction rule for
%a GADT can be derived as an instance of its deep induction rule.
But it actually delivers far more: it gives a {\em general framework}
for deriving deep induction rules for GADTs that can be instantiated
to particular such data type of interest. This framework can serve as
a basis for extending modern proof assistants' automatic generation of
structural induction rules for ADTs to automatic generation of deep
induction rules for GADTs. In addition, as for ADTs and nested types,
the structural induction rule for any GADT can be recovered from its
deep induction rule simply by taking the custom predicates in its deep
induction rule to be constantly $\mathsf{True}$-valued (i.e.,
constantly $\mathsf{\top}$-valued) predicates.

Significantly, deep induction rules for GADTs cannot be derived by
somehow extending the approach of~\cite{jp20} to syntax-only
GADTs. Indeed, the approach taken there makes crucial use of the
functoriality of data types' interpretations from~\cite{jp19}, and
functoriality is precisely what interpreting GADTs as the
interpretations of their Church encodings fails to deliver. Our
approach is to instead first give a predicate lifting styled after
those of~\cite{jp20}, together with a (deep) induction rule, for the
simplest --- and arguably most important --- GADT, namely the equality
GADT~\eqref{eq:equal}. We then derive the deep induction rule for a
more complex GADT $\mathsf{G}$ by {\em i}) using the equality GADT to
represent $\mathsf{G}$ as its so-called {\em Henry Ford
  encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii}) using
the predicate liftings for the equality GADT and the other GADTs
appearing in the definition of $\mathsf{G}$ to appropriately thread
the custom predicates for the primitive types appearing in
$\mathsf{G}$ throughout $\mathsf{G}$'s structure. This two-step
process delivers deep induction rules for a very general class of
GADTs. To illustrate, we introduce a series of increasingly complex
GADTs as running examples in Section~\ref{sec:GADTs} and derive a deep
induction rule for each of them in
Section~\ref{sec:deep-ind-GADTs}. In particular, we derive the deep
induction rules for the equality data type in
Section~\ref{sec:ind-equal} and the $\mathsf{Seq}$ data type
in~\eqref{eq:seq} in Section~\ref{sec:ind-seq}. We present our general
framework for deriving (deep) induction rules for GADTs in
Section~\ref{sec:framework}, and observe that the derivations in
Section~\ref{sec:deep-ind-GADTs} are all instances of it.  In
Section~\ref{sec:GADT-nested} we show that, by contrast with truly
nested types, which do have a functorial semantics, syntax-only GADTs'
lack of functoriality means that it is not clear how to extend
induction --- deep or otherwise --- to {\em truly nested GADTs}, i.e.,
to proper GADTs whose recursive occurrences appear below
themselves.\footnote{Note carefully the distinction between a GADT
  that is not a nested type --- i.e., a proper GADT --- and a proper
  GADT that is not a truly nested GADT.  In fact, truly nested types
  are not proper GADTs and truly nested GADTs are not (truly) nested
  types. There is ample scope for confusion in light of the original,
  and now well-established, use of the term ``nested type'' to refer
  to {\em any} type that allows non-variable instances in the domains
  of its constructors, whether or not that type involves actual
%  types that do not necessarily involve actual
  nesting~\cite{bm98}.}  This does not appear to be much of a
restriction, however, since truly nested GADTs do not, to our
knowledge, appear in practice or in the
literature. Section~\ref{sec:app} comprises a case study in using deep
induction. All of the deep induction rules appearing in this paper
have been derived by instantiating our general framework. Our Agda
code implementing them is available at~\cite{web-page}.

@@

{\bf Additional Related Work\/} Various techniques for deriving
induction rules for data types that go beyond ADTs have been
studied. For example, Fu and Selinger~\cite{fs18} show, via examples,
how to derive induction rules for arbitrary nested
types. Unfortunately, however, their technique is rather {\em ad hoc},
so is unclear how to generalize it to nested types other than the
specific ones studied there. Moreover,~\cite{fs18} actually derives
induction rules for data types {\em related} to the original nested
types rather than for the original nested types themselves, and it is
unclear whether or not the derived rules are sufficiently expressive
to prove all results about the original nested types that we would
expect to be provable by induction. This latter point echoes the issue
with Coq-derived induction rule for rose trees raised in
Section~\ref{sec:intro}, which has the unfortunate effect of forcing
users to manually write induction (and other) rules for such types for
use in that system. Tassi~\cite{tas19} derives induction rules for
data type definitions in Coq using unary parametricity. His technique
seems to be essentially equivalent to that of~\cite{jp19} for nested
types, although he does not permit true nesting. More recently,
Ullrich~\cite{ull20} has implemented a plugin in MetaCoq to generate
induction rules for nested types. This plugin is also based on unary
parametricity, and true nesting still is not permitted.  As far as we
know, no attempts have been made to extend either implementation to
truly nested types or to proper GADTs or their deep variants. In fact,
we know of no work other than that reported here that specifically
addresses induction rules for these data types.

@@

Under some circumstances deep induction can be mimicked by
hand-threading applications of structural induction through the layers
of data comprising a deep data type. But this is not the case if,
e.g., one or more of the custom predicates in the data type's deep
induction rule is not the characteristic function of an inductive data
type.
%cannot be defined by structural induction.

@@

GADTs are used in precisely those situations in which different
behaviors at different instances of data types are desired. This is
achieved by allowing the programmer to give the type signatures of the
GADT's data constructors independently, and then taking advantage of
pattern matching to force the desired type refinement. For example,
the {\em equality} GADT
\begin{equation}\label{eq:equal}
\begin{array}{l}
\mathsf{data\ Equal : Set \to Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,refl :\, \forall \{A : Set\} \to Equal\,A\,A}
\end{array}
\end{equation}
is parameterized by two type indices, but it is only possible to
construct data elements of type $\mathsf{Equal\,A\,B}$ if $\mathsf{A}$
and $\mathsf{B}$ are instantiated at the same type. If the types
$\mathsf{A}$ and $\mathsf{B}$ are {\color{red} propositionally equal}
%syntactically identical
then the
type $\mathsf{Equal\,A\,B}$ contains the single data element
$\mathsf{refl}$. It contains no data elements otherwise.

@@

The equality constraints engendered by GADTs' data constructors makes
deriving (deep) induction rules for them more involved than for ADTs
and other nested types. Nevertheless, we show in this section how to
do so. We first illustrate the key components of our approach by
deriving deep induction rules for the three specific GADTs introduced
in Section~\ref{sec:GADTs}. Then, in Section~\ref{sec:framework}, we
abstract these to a general framework that can be applied to any GADT
that is not a truly nested GADT. As hinted above, the predicate
lifting for the equality GADT plays a central role in deriving both
structural and deep induction rules for more general GADTs.

@@

To define the (deep) induction rule for any GADT $\mathsf{G}$ we first
need to define a predicate lifting that maps a predicate on a type
$\mathsf{A}$ to a predicate on $\mathsf{G\,A}$. Such a predicate
lifting $\mathsf{Equal^{\wedge} : \forall (A\,B : Set) \to (A \to Set)
  \to (B \to Set) \to Equal\,A\,B \to Set}$ for $\mathsf{Equal}$ is
defined by $\mathsf{Equal^{\wedge}\,A\,A\,Q\,Q'\,refl = \forall (a :
  A) \to Equal\,(Q\,a)(Q'\,a)}$.  It takes two predicates on the same
type as input and tests them for extensional equality.
%
Next, we need to associate with each data constructor $\mathsf{c}$ of
$\mathsf{G}$ an {\em induction hypothesis} asserting that, if the
custom predicate arguments to a predicate $\mathsf{P}$ on $\mathsf{G}$
can be lifted to $\mathsf{G}$ itself, then $\mathsf{c}$ {\em respects}
$\mathsf{P}$, i.e., $\mathsf{c}$ constructs data satisfying the
instance of $\mathsf{P}$ at those custom predicates. The following
induction hypothesis $\mathsf{dIndRefl}$ is thus associated with the
$\mathsf{refl}$ constructor for $\mathsf{Equal}$:
\begin{equation*}\label{eq:ind-refl}
\begin{array}{l}
\mathsf{\lambda (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
  Set) \to Equal\,A\,B \to Set) \to} \\ 
\quad\mathsf{\forall (C : Set) (Q\, Q' : C \to Set) \to
  Equal^{\wedge}\,C\,C\,Q\,Q'\,refl \to P\,C\,C\,Q\,Q'\,refl} 
\end{array}
\end{equation*}
The deep induction rule for $\mathsf{G}$ now states that, if all of
$\mathsf{G}$'s data constructors respect a predicate $\mathsf{P}$,
then $\mathsf{P}$ is satisfied by every element of $\mathsf{G}$ to
which the custom predicate arguments to $\mathsf{P}$ can be
successfully lifted.  The deep induction rule for $\mathsf{Equal}$ is
thus
\begin{equation}\label{eq:ind-equal}
\begin{array}{l}
\mathsf{\forall (P : \forall (A\,B : Set) \to (A \to Set) \to (B \to
  Set) \to Equal\,A\,B \to Set) \to dIndRefl\,P \to}\\ \quad 
  \mathsf{\forall (A\,B : Set) (Q_A : A \to Set) (Q_B : B \to Set) (e:
  Equal\,A\,B) \to Equal^{\wedge}\,A\,B\,Q_A\,Q_B\,e \to
  P\,A\,B\,Q_A\,Q_B\,e}
\end{array}
\end{equation}

@@

We can generalize the approach taken in
Section~\ref{sec:deep-ind-GADTs} to a general framework for deriving
deep induction rules for GADTs that are not truly nested GADTs. We
will treat GADTs of the form

@@

\noindent
This grammar is subject to the following restrictions. In the third
clause the type constructor $\mathsf{F_1}$ does not contain
$\mathsf{G}$. In the fourth clause, none of the $\mathsf{\alpha}$-many
type constructors in $\mathsf{\ol{F_1}}$ contains $\mathsf{G}$. This
prevents nesting, which would {\color{red} make it impossible to give}
an induction rule for $\mathsf{G}$; see Section~\ref{sec:GADT-nested}
below. In the fifth and sixth clauses, $\mathsf{H : Set^\gamma \to
  Set}$ is the syntactic reflection of some functor, and thus has an
associated map function. It is worth noting that the fifth clause
subsumes the cases in which $\mathsf{F\,G\,\ol{B}}$ is a closed type
or one of the $\mathsf{B_i}$, and that $\mathsf{H}$ can be the data
type constructor for any (truly) nested type. From the map function
for $\mathsf{H}$ we can also construct a map function
